---
title: "Multilevel Modelling: Assignment 2"
author: "Ilaria Lunardelli, Giuliana Orizzonte, Kyra Giran"
date: "28/03/2023"
output: pdf_document
---
```{r, results= 'asis', message=F, warning=F, echo=F}
# Libraries: add libraries used here
library(lme4)
library(lattice)
library(lmerTest)
library(ggplot2)
library(tidyverse)
library(texreg)
library(readr)
library(tidyr)
library(gtsummary)
library(jtools)

```

```{r, echo=F, results='hide'}
# Load Data
curran <- read_csv("curran_wide.csv")
#Summary Statistics 
summary(curran)
# this gives us the full number to copy into the excel sheet
options(scipen=999)
```

#### Question 1

Convert the wide data file into a long format. Check the data and recode if necessary.
• Check the linearity assumption, report and include plots.
• Check for outliers (don’t perform analyses, just look in the scatterplots), report.

Looking at the scatterplots we see no clear indication that the data would be modeled by a quadratic function better than a linear one. Therefore, we assume that the relation between predictors and the dependent variable can be modeled linearly.

Looking at the scatterplots, we can see the presence of outliers, in the upper tail of the antisocial behavior measurement scale.

```{r, echo=F, warning=FALSE, message=FALSE, results = 'hide'}
# Convert the data to long format
curran_long <- pivot_longer(data = curran, cols = c(2:9),                   
                         names_to = c(".value", "time"),                   
                         names_pattern = "(anti|read)(.)")
head(curran_long)       # check dataset head
# Convert time to factor for graphs
curran_long$time <- as.factor(curran_long$time)
#grand mean centering
curran_long$read_gmc <- curran_long$read - mean(curran_long$read)
curran_long$momage_gmc <- curran_long$momage - mean(curran_long$momage)
curran_long$homecog_gmc <- curran_long$homecog - mean(curran_long$homecog)
# Plots to check the linearity assumption
ggplot(curran_long,
       aes(x = time, y = anti)) +
  geom_point(color="#666666") +
  geom_smooth(color= "#6699FF", method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(color= "#FF3333", method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) +
  theme_bw() +
  labs( title = "Scatterplot Time-Anti")
ggplot(curran_long,
       aes(x = read_gmc, y = anti)) +
  geom_point(color="#666666") +
  geom_smooth(color= "#6699FF", method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(color= "#FF3333", method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) +
  theme_bw() +
  labs( title = "Scatterplot Read-Anti")
# Boxplot with outliers
#ggplot(curran_long,
 #      aes(x=time, y=anti, group=time, color=time)) + 
#  geom_boxplot() +
 # scale_x_discrete(breaks = c(0,1,2,3)) + 
  #theme(legend.position = 'none', text = element_text(size = 7)) +
  #labs(
   # title = 'Graph 3: Boxplot with median, 1st and 3rd Quartiles, IQR, and outliers of antisocial scores by measurement wave',
    #x = 'Measurement Wave',
    #y = 'Antisocial Score'
  #)
```

#### Question 2

Answer the question: should you perform a multilevel analysis?
• What is the mixed model equation?
• Provide and interpret the relevant results.
• What is the intraclass correlation?
• What is your conclusion regarding the overall question regarding the necessity of performing a multilevel analysis?

We should perform a multilevel analysis because we can expect measurements from the same individual to be much more similar than measurements across different individuals. Multilevel analysis accounts for this similarity and keeps the Type I error at .05.

###### Mixed Model Equations
These are the model equations a priori, meaning they include all predictors and possible interactions. However, during the analysis, we remove any predictors that result non-significant.

Model for calculating ICC:
- $Antisocial_{ti} = \beta_{00} + u_{0i} + \epsilon_{ti}$

Baseline model (time is fixed): 
- $Antisocial_{ti} = \beta_{00} + \beta_{10}Time_{ti} + u_{0i} + \epsilon_{ti}$

With time-varying covariates to predict within person variability:
- $Antisocial_{ti} = \beta_{00} + \beta_{10}Time_{ti} + \beta_{20}Reading_{ti} + u_{0i} + \epsilon_{ti}$

With time-invariant covariates to predict between persons variability:
- $Antisocial_{ti} = \beta_{00} + \beta_{10}Time_{ti} + \beta_{20}Reading_{ti} + \beta_{01}MotherAge_{i} + \beta_{02}CognitiveStimulation_{i} + u_{0i} + \epsilon_{ti}$

Allow for random time-effect:
- $Antisocial_{ti} = \beta_{00} + \beta_{10}Time_{ti} + \beta_{20}Reading_{ti} + \beta_{01}MotherAge_{i} + \beta_{02}CognitiveStimulation_{i} + u_{1i}Time_{ti} + u_{0i} + \epsilon_{ti}$

With cross level interaction to predict slope variability:
- $Antisocial_{ti} = \beta_{00} + \beta_{10}Time_{ti} + \beta_{20}Reading_{ti} + \beta_{01}MotherAge_{i} + \beta_{02}CognitiveStimulation_{i} + \beta_{11}Time_{ti}MotherAge_{i} + \beta_{12}Time_{ti}CognitiveStimulation_{i} + \beta_{21}Reading_{ti}MotherAge_{i} + \beta_{22}Reading_{ti}CognitiveStimulation_{i} + u_{1i}Time_{ti} + u_{0i} + \epsilon_{ti}$

###### The ICC

The intraclass correlation is 0.48, which represents the expected correlation between two random time points for the same child, as well as the percentage of variance at the child level (time invariant). The ICC high enough to require multilevel analysis in order to not incur in a spurious significant result.

###### Results and Interpretation

```{r, echo=F, results='hide', warning=F, message=F}
# Center time so that the first occurrence is 0
curran_long$time <- as.numeric(curran_long$time) - 1
mod0 <- lm(anti~1, data = curran_long)
summary(mod0)
var(resid(mod0))

mod1 <- lmer(anti ~ 1+ (1|id), REML = F, data=curran_long)
summary(mod1)
performance::icc(mod1)
anova(mod1, mod0)
```
Model 0 ignores the multilevel structure. It gives us an intercept ($\beta_{00}$ = 1.82, $SE$ = 0.06) which represents the mean `Antisocial` behaviour level.

The AIC is 3573.5.

Model 1 performs a multilevel analysis without any predictors.

The intercept ($\beta_{00}$ = 1.82, $SE$ = 0.10) represents mean of `Antisocial` behaviour across people.

The `Time` variance ($\u_{0i}$ = 1.58, $SE$ = 1.26) is the random effect capturing the variance of people around their own average antisocial behaviour level and the child variance ($\\epsilon_{ti}$ = 1.74, $SE$ = 1.32) is the random effect for the intercept capturing the variance of people’s average antisocial behaviour levels around the intercept.

The comparison of model 0 and model 1 is significant ($\chi^2$(1) = 231.97/2 = 115.98, (*p* < .001)), which indicates that the two differ in how well they fit the data. The AIC of model 1 is 3343.5, which is lower than that of model 0, and thus indicates a better fit.

```{r, echo=F, results='hide', warning=F, message=F}
mod2 <- lmer(anti ~ 1 + time + (1|id), REML = F, data=curran_long)
summary(mod2)
anova(mod1, mod2)
```

Model 2 includes only `Time` as a fixed effect at level one.

The intercept ($\beta_{00}$ = 1.55, $SE$ = 0.11) represents the mean of `Antisocial` behaviour across people. The fixed effect for `Time` is significant ($\beta_{10}$ = 0.18, SE = 0.04, t(663) = 4.51, *p* < .001) and it represents the average increase in `Antisocial` behaviour levels over one unit of time.

The `Time` variance ($\u_{0i}$ = 1.69, $SE$ = 1.30) is the random effect capturing the variance of children around their own average antisocial behaviour level, whereas the child variance ($\epsilon_{ti}$ = 1.59, $SE$ = 1.26) is the random effect for the intercept capturing the variance of the children’s average antisocial behaviour levels around the intercept.

The comparison of model 1 and model 2 is significant ($\chi^2$(1) = 20.06/2 = 10.03, *p* < .001), which indicates that the two differ in how well they fit the data. The AIC of model 1 is 3325.5, which is lower than that of model 0, and thus indicates a better fit.

#### Question 3

Add the time-varying predictor(s).
• Provide and interpret the relevant results and provide your overall conclusion.

```{r, echo=F, results='hide', warning=F, message=F}
mod3 <- lmer(anti ~ 1 + time + read_gmc + (1|id), REML = F, data=curran_long)
summary(mod3)
anova(mod2, mod3)
```

Model 3 includes as predictors the measurement waves `Time` and the children `Reading` recognition score as fixed effects at level one, representing the time-varying effects.

The intercept ($\beta$ = 1.50, SE = .15) represents the mean of `Antisocial` behaviour across people. `Time` is a significant predictor ($\beta_{10}$ = 0.21, SE = .08, t(882) = 2.73, *p* = .006) and it represents the average increase in `Antisocial` behaviour levels over one unit of time. The `Reading` recognition predictor, which represents the effect of reading recognition at time = 0, is not significant ($\beta_{20}$ = -0.03, SE = .06, t(830) = -0.54, *p* = .588). Thus, we conclude that the level of `Antisocial` behavior tends to increase with time and is not affected by reading skills.

The time variance ($\u_{0i}$ = 1.69, $SE$ = 1.30) is the random effect capturing the variance of people around their own average antisocial behaviour level, whereas the child variance ($\epsilon_{ti}$ = 1.58, $SE$ = 1.26) is the random effect for the intercept capturing the variance of the children’s average antisocial behaviour levels around the intercept.

The comparison of model 2 and model 3 is not significant ($\chi^2$(1) = 0.29/2=0.15, *p* = 0.59), which indicates that the two fit the data similarly. However, the AIC of model 1 is 3327.2, which is higher than that of the previous model, thus suggesting a worse model fit. This is likely due to the fact that `Reading` is not a useful predictor. Therefore, `Reading` is removed from subsequent models.

#### Question 4

On which level or levels can you expect explained variance?
• Calculate and interpret the explained variances

The model returns explained variances for both level 1 ($\sigma_{time}$ = 1.69, $sd$ = 1.30) and level 2 ($\sigma_{child}$ = 1.58, $sd$ = 1.26). The explained variances indicates how much variation in `Antisocial` behavior is explained by our predictors. In other words, they are part of the variation in antisocial behavior that is explicitly modeled and accounted for by our predictors.

Using model 2 as baseline, we can calculate R2 for both level 1 ($\R^2_{level1}$ = -0.002) and level 2 ($\R^2_{level2}$ = -0.01), which measure the explained variance at the time level and at the individual level.

#### Question 5

Add the time invariant predictor(s) to the model.
• Provide and interpret the relevant results and provide your overall conclusion.

```{r, echo=F, results='hide', warning=F, message=F}
mod4 <- lmer(anti ~ 1 + time + momage_gmc + homecog_gmc + (1|id), REML = F, data=curran_long)
summary(mod4)
anova(mod3, mod4)
```

Model 4 includes `Time` as the fixed effects at level one and `Mom Age` as well as `Home Cognitive Stimulation` as fixed effects at level two, which is at the children level. 

The results show that `Time` ($\beta$ = 0.18, SE = .04, t(663) = 4.51, *p* < .001) and `Home Cognitive Stimulation` ($\beta$ = -0.13, SE = .04, t(220) = -3.35, *p* < .001) are, respectively, a positive and a negative predictor of `Antisocial` behavior. Conversely, `Mom Age` seems to have no effect ($\beta$ = 0.00, SE = .05, t(220) = -0.02, *p* = .985). In other words, antisocial behavior appears to increase over time and decrease with higher levels of cognitive stimulation, while the mother's age at birth is not an influential factor.

Since `Mom Age` is not a significant predictor, it is removed from subsequent models.

The comparison of model 3 and model 4 is significant ($\chi^2$(1) = 11.36/2=5.68, *p* < .001), which indicates that the two differ in how well they fit the data. The AIC of model 4 is 3317.8, which is lower than that of model 3, and thus indicates a better fit.

#### Question 6

On which level or levels can you expect explained variance?
• Calculate and interpret the explained variances.

Similarly to model 3, model 4 returns explained variances for both level 1 ($\sigma_{time}$ = 1.69, $sd$ = 1.30) and level 2 ($\sigma_{child}$ = 1.49, $sd$ = 1.22), with the difference that now predictors at the children level are also included. The explained variance at level 1 did not change because we did not change the time-varying predictors. Meanwhile, the explained variance at level 2 slightly decreased, meaning that we are able to account for slightly less of the variance in antisocial behavior.

At this stage, we will still have the same R squared for level 1 ($\R^2_{level1}$ = -0.002) since we didn't change predictors for level 1 but, R squared at level 2 ($\R^2_{level2}$ = -0.01) increased, meaning that the variance accounted by level 2 predictors increased.

#### Question 7

For the time-varying predictor(s), check if the slope is fixed or random.
• What are the null- and alternative hypotheses?
• Provide and interpret the relevant results.
• Provide an overall conclusion

The null hypothesis is that there is no variation in the slopes of the regression lines across different children, and that any slope differences observed between children are due to chance. 

The alternative hypothesis is that there is significant between-subjects variation in the slopes, meaning that the relation between the predictors and antisocial behavior varies across different children.

```{r, echo=F, results='hide', warning=F, message=F}
mod5a <- lmer(anti ~ 1 + time + homecog_gmc + (1 + time|id), REML = F, data=curran_long)
summary(mod5a)
anova(mod4, mod5a)

mod5b <- lmer(anti ~ 1 + time + read_gmc + homecog_gmc + (1 + read_gmc|id), REML = F, data=curran_long)
summary(mod5b)
anova(mod4, mod5b)
```

In Model 5 the slope of `Time` is allowed to vary between children.

The significant level 1 predictor `Time` ($\beta$ = 0.18, SE = .04, t(221) = 4.14, *p* < .001) suggests that `Antisocial` behavior on average tends to increase over further assessments. The significant level 2 predictor `Home Cognitive Stimulation` ($\beta$ = -0.10, SE = .04, t(220) = -2.79, *p* = .006) suggests that higher cognitive stimulation results in less `Antisocial` behavior. The `Time` random effect suggests that children differ in how their `Antisocial` behavior changes over measurement occasions.

The comparison of model 5 and model 4 is significant ($\chi^2$(1) = 26.56/2=13.28, *p* < .001), which indicates that the two differ in how well they fit the data. The AIC of model 5a is 3293.3, which is lower than that of model 4, and thus indicates a better fit.

Overall, antisocial behavior appears to increase over time and decrease with higher levels of cognitive stimulation, however the rate of change varies for individual children.

We considered testing a similar model allowing random slopes for the `Reading` predictor; however, this predictor has previously been shown to be non-significant, meaning that there should be no effect of `Reading` on `Antisocial` behavior. Therefore, we found it unnecessary to test whether the change in `Antisocial` behavior varies at different `Reading` scores, because we previously concluded the reading scores may not change antisocial behavior at all. Therefore, said model is not part of our analysis.

#### Question 8

If there is a random slope, set up a model that predicts the slope variation.
• Provide and interpret the relevant results and provide your overall conclusion.

```{r}
mod6a<- lmer(anti ~ 1 + time + homecog_gmc + read_gmc + momage_gmc + time*homecog_gmc + time*momage_gmc + read_gmc*momage_gmc + read_gmc*homecog_gmc + (time|id), REML = F, data=curran_long)
summary(mod6a)
```
```{r, echo=F, results='hide', warning=F, message=F}
mod6b <- lmer(anti ~ 1 + time + homecog_gmc + time*homecog_gmc + (time|id), REML = F, data=curran_long)
summary(mod6b)
anova(mod5a, mod6b)
```

In Model 6a we reintroduce all of the predictors to check whether non-significant predictors are still part of significant cross-level interactions.

From the results we see that this is not the case. In this model, none of the predictors or interactions are significant (0.18 < *p* < .99), with the exception of `Time` ($\beta$ = 0.18, SE = .08, t(662) = 2.14, *p* = .03).

Given that this model appears to be overall a poor predictor of antisocial behavior we move on to a model that includes only previously significant predictors and their cross-level interactions.

Model 6b we reintroduce all of the predictors to check whether non-significant predictors are still part of significant cross-level interactions.

The significant level 1 predictor `Time` ($\beta$ = 0.18, SE = .04, t(221) = 4.20, *p* < .001) suggests that `Antisocial` behavior on average tends to increase over further assessments. The level 2 predictor `Home Cognitive Stimulation` ($\beta$ = -0.06, SE = .02, t(221) = -1.63, *p* = .11) is no longer significant, while the interaction `Time x Home Cognitive Stimulation` becomes significant instead ($\beta$ = -0.04, SE = .02, t(221) = -2.64, *p* = .009). 

The comparison of model 6b and model 5a is significant ($\chi^2$(1) = 6.88/2 = 3.44, *p* = .009), which indicates that the two differ in how well they fit the data. The AIC of model 6b is 3288.4, which is lower than that of model 4, and thus indicates a better fit.

Overall, we see that antisocial behavior increases with time and the effect of cognitive stimulation on antisocial behavior depends on the passing of time. Because the direct effect of cognitive stimulation is negative and the interaction between time and cognitive stimulation is also negative, it seems that the effect of cognitive stimulation becomes more pronounced as time goes on.

#### Question 9

Decide on a final model.
• provide the separate level 1 and 2 model equations, as well as the mixed model equation.
• Check the normality assumption for both the level-1 and level-2 errors, report

###### Model Equations
- Overall:
$Antisocial_{ti} = \beta_{00} + \beta_{10}Time_{ti} + \beta_{01}CognitiveStimulation_{i} + \beta_{11}Time_{ti}CognitiveStimulation_{i} + u_{1i}Time_{ti} + u_{0i} + \epsilon_{ti}$

$Antisocial_{ti} = 1.55 + 0.18 * Time_{ti} + -0.06 * CognitiveStimulation_{i} + -0.05 * Time_{ti} * CognitiveStimulation_{i} + 0.08 * Time_{ti} + 0.94 + 1.53$

- Time level:
$Antisocial_{ti} = \pi_{0i} + \pi_{1i}Time_{ti} + \epsilon_{ti}$

$Antisocial_{ti} = \pi_{0i} + \pi_{1i}Time_{ti} + 1.53$

- Children level:
$pi_{0i} = \beta_{00} + \beta_{01}CognitiveStimulation_{i} + u_{0i}$
$pi_{1i} = \beta_{10} + \beta_{11}CognitiveStimulation_{i} + u_{1i}$

$pi_{0i} = 1.55 + -0.06 * CognitiveStimulation_{i} + 0.94$
$pi_{1i} = 0.18 + -0.05 * CognitiveStimulation_{i} + 0.08$

- Normality assumption

```{r, echo=F, results='hide', warning=F, message=F}
#first level
qqnorm(residuals(mod6b), col= "#FF3333", main= "Level 1")

#second level
#intercept
qqnorm(ranef(mod6b)$id[,1], col= "#6699FF", main= "Level 2, intercept")

#slope
qqnorm(ranef(mod6b)$id[,2], col= "#00FFCC", main= "Level 2, slope")
```

From the qq-plots we can check the normality assumption for level 1 and level 2 residuals. At level 1, we can say that the distribution of the residuals is roughly normal, however, at the second level, both intercept and slope have right skewed distributions of residuals. As a consequence, the normality assumption holds for level 1 residuals but doesn not for level 2 ones.


#### Contributions: 
Methods and Analysis: Ilaria Francesca Lunardelli , Giuliana Orizzonte and Kyra Giran; Visualisations: Ilaria Francesca Lunardelli , Giuliana Orizzonte and Kyra Giran; Model interpretation: Ilaria Francesca Lunardelli , Giuliana Orizzonte and Kyra Giran; Proof and Editing: Ilaria Francesca Lunardelli , Giuliana Orizzonte and Kyra Giran.







