---
title: "Multilevel Modelling: Assignment 2"
author: "Ilaria Lunardelli, Giuliana Orizzonte, Kyra Giran"
date: "28/03/2023"
output: pdf_document
---
```{r, message=F, warning=F, echo=F}
# Libraries: add libraries used here
library(lme4)
library(lattice)
library(lmerTest)
library(ggplot2)
library(tidyverse)
library(texreg)
library(readr)
library(tidyr)
library(gtsummary)
```
```{r, echo=F, results='hide'}
# Load Data
curran <- read_csv("curran_wide.csv")

#Summary Statistics 
summary(curran)

# this gives us the full number to copy into the excel sheet
options(scipen=999)   

# Center time so that the first occurrence is 0
curran_long$time <- as.numeric(curran_long$time) - 1

# Convert time to factor
curran_long$time <- as.factor(curran_long$time)

#grand mean centering reading score 
curran_long$read_gmc <- curran_long$read - mean(curran_long$read)

# grand mean centering 
curran_long$momage_gmc <- curran_long$momage - mean(curran_long$momage)
curran_long$homecog_gmc <- curran_long$homecog - mean(curran_long$homecog)

```

#### Question 1

Convert the wide data file into a long format. Check the data and recode if necessary.
• Check the linearity assumption, report and include plots.
• Check for outliers (don’t perform analyses, just look in the scatterplots), report.

```{r, echo=F, results = 'hide'}

# Convert the data to long format
curran_long <- pivot_longer(data = curran, cols = c(2:9),                   
                         names_to = c(".value", "time"),                   
                         names_pattern = "(anti|read)(.)")
head(curran_long)       # check dataset head

# Plots to check the linearity assumption
ggplot(curran_long,
       aes(x = time, y = anti)) +
  geom_point() +
  geom_smooth(method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) 

ggplot(curran_long,
       aes(x = anti, y = read)) +
  geom_point() +
  geom_smooth(method = "lm",
              aes(color = "linear"),
              se = FALSE) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              aes(color = "quadratic"),
              se = FALSE) 

# Boxplot with outliers
ggplot(curran_long,
       aes(x=time, y=anti, group=time, color=time)) + 
  geom_boxplot() +
  scale_x_discrete(breaks = c(0,1,2,3)) + 
  theme(legend.position = 'none', text = element_text(size = 7)) +
  labs(
    title = 'Graph 3: Boxplot with median, 1st and 3rd Quartiles, IQR, and outliers of antisocial scores by measurement wave',
    x = 'Measurement Wave',
    y = 'Antisocial Score'
  )
```

#### Question 2

Answer the question: should you perform a multilevel analysis?
• What is the mixed model equation?
• Provide and interpret the relevant results.
• What is the intraclass correlation?
• What is your conclusion regarding the overall question regarding the necessity of performing a multilevel analysis?

yti = β00 + u0i + π1iTti + eti

Model for calculating ICC:
- $Antisocial_{ti} = \pi_{00} + u_{0i} + \epsilon_{ti}$

Baseline model (time is fixed): 
- $Antisocial_{ti} = \pi_{00} + \pi_{10}Time_{ti} + u_{0i} + \epsilon_{ti}$

With time-varying covariates to predict within person variability:
- $Antisocial_{ti} = \pi_{00} + \pi_{10}Time_{ti} + \pi_{20}Reading_{ti} + u_{0i} + \epsilon_{ti}$

With time-invariant covariates to predict between persons variability:
- $Antisocial_{ti} = \pi_{00} + \pi_{10}Time_{ti} + \pi_{20}Reading_{ti} + \beta_{01}MotherAge_{i} + \beta_{02}CognitiveStimulation_{i} + u_{0i} + \epsilon_{ti}$

Allow for random time-effect:
- $Antisocial_{ti} = \pi_{00} + \pi_{10}Time_{ti} + \pi_{20}Reading_{ti} + \beta_{01}MotherAge_{i} + \beta_{02}CognitiveStimulation_{i} + u_{1i}Time_{ti} + u_{0i} + \epsilon_{ti}$

With cross level interaction to predict slope variability:
- $Antisocial_{ti} = \pi_{00} + \pi_{10}Time_{ti} + \pi_{20}Reading_{ti} + \beta_{01}MotherAge_{i} + \beta_{02}CognitiveStimulation_{i} + \beta_{21}MotherAge_{i}Reading_{ti} + \beta_{22}CognitiveStimulation_{i}Reading_{ti} + u_{1i}Time_{ti} + u_{0i} + \epsilon_{ti}$

```{r, echo=F, results='hide', warning=F, message=F}
# ICC Calculations 

mod1 <- lmer(anti ~ 1+ (1|id), REML = F, data=curran_long)
summary(mod1)
performance::icc(mod1)

```
```{r, echo=F, results='hide', warning=F, message=F}

mod2 <- lmer(anti ~ 1 + time + (1|id), REML = F, data=curran_long)
summary(mod2)
performance::icc(mod2)

```
#### Question 3

Add the time-varying predictor(s).
• Provide and interpret the relevant results and provide your overall conclusion.

#### Question 4

On which level or levels can you expect explained variance?
• Calculate and interpret the explained variances

```{r, echo=F, results='hide', warning=F, message=F}

mod3 <- lmer(anti ~ 1 + time + read_gmc + (1|id), REML = F, data=curran_long)
summary(mod3)
performance::icc(mod3)

```

#### Question 5

Add the time invariant predictor(s) to the model.
• Provide and interpret the relevant results and provide your overall conclusion.

```{r, echo=F, results='hide', warning=F, message=F}

mod4 <- lmer(anti ~ 1 + time + read_gmc + momage_gmc + homecog_gmc + (1|id), REML = F, data=curran_long)
summary(mod4)
performance::icc(mod4)

```

#### Question 6

On which level or levels can you expect explained variance?
• Calculate and interpret the explained variances.

#### Question 7

For the time-varying predictor(s), check if the slope is fixed or random.
• What are the null- and alternative hypotheses?
• Provide and interpret the relevant results.
• Provide an overall conclusion

```{r, echo=F, results='hide', warning=F, message=F}

mod5a <- lmer(anti ~ 1 + time + read_gmc + momage_gmc + homecog_gmc + (1 + time|id), REML = F, data=curran_long)
summary(mod5a)
performance::icc(mod5a)

```
```{r, echo=F, results='hide', warning=F, message=F}

mod5b <- lmer(anti ~ 1 + time + read_gmc + momage_gmc + homecog_gmc + (1 + read|id), REML = F, data=curran_long)
summary(mod5b)
performance::icc(mod5b)

```

#### Question 8

If there is a random slope, set up a model that predicts the slope variation.
• Provide and interpret the relevant results and provide your overall conclusion.

```{r, echo=F, results='hide', warning=F, message=F}

mod6 <- lmer(anti ~ 1 + time + read_gmc + momage_gmc + homecog_gmc + time*read_gmc+ (time|id), REML = F, data=curran_long)
summary(mod6)
performance::icc(mod6)

```

#### Question 9

Decide on a final model.
• provide the separate level 1 and 2 model equations, as well as the mixed model equation.
• Check the normality assumption for both the level-1 and level-2 errors, report


#### Contributions:\
\